---
title: "Analysis of Temperature and Rainfall Trends Using R"
author: "R Ambarish"
date: "`r format(Sys.Date(), '%d-%m-%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
---

# 1. Introduction

Weather conditions change from day to day. Temperature and rainfall are two important variables used to understand these changes. This Tiny Project uses **R and R Markdown** to import, preprocess, analyse and visualize a weather dataset.

**Dataset note:** `weather_data.csv` is a self-contained 30-day sample dataset prepared for academic demonstration.

# 2. Objectives

- Import weather data from a CSV file.
- Inspect, preprocess and clean the data.
- Calculate descriptive statistics.
- Study temperature and rainfall trends.
- Create graphs and plots.
- Examine the relationship between temperature and rainfall.
- Interpret the results.
- Present findings and a conclusion.

# 3. Data Importing

The data is stored in a CSV file. The `read.csv()` function imports it into R.

```{r import}
weather <- read.csv("weather_data.csv", stringsAsFactors = FALSE)
weather$Date <- as.Date(weather$Date)

head(weather)
str(weather)
```

**Code explanation:** `read.csv()` reads the CSV file. `as.Date()` converts the Date column into a proper date format. `head()` displays the first rows and `str()` shows the structure of the dataset.

# 4. Data Preprocessing and Cleaning

First, missing values and duplicate rows are checked.

```{r checks}
colSums(is.na(weather))
sum(duplicated(weather))
```

The data is then cleaned by removing duplicate records and incomplete rows.

```{r cleaning}
weather <- weather[!duplicated(weather), ]
weather <- weather[complete.cases(weather), ]

nrow(weather)
```

**Code explanation:** `duplicated()` identifies repeated rows. `complete.cases()` identifies complete records. The filtering statements keep only unique and complete observations.

# 5. Statistical Analysis

The following code calculates mean, median, minimum, maximum and standard deviation.

```{r statistics}
summary_table <- data.frame(
  Measure = c("Mean", "Median", "Minimum", "Maximum", "Standard Deviation"),
  Temperature_C = c(
    mean(weather$Temperature_C),
    median(weather$Temperature_C),
    min(weather$Temperature_C),
    max(weather$Temperature_C),
    sd(weather$Temperature_C)
  ),
  Rainfall_mm = c(
    mean(weather$Rainfall_mm),
    median(weather$Rainfall_mm),
    min(weather$Rainfall_mm),
    max(weather$Rainfall_mm),
    sd(weather$Rainfall_mm)
  )
)

knitr::kable(round(summary_table, 2),
             caption = "Summary Statistics")
```

**Code explanation:** `mean()` calculates the average, `median()` gives the middle value, `min()` and `max()` find the lowest and highest values, and `sd()` measures variation.

# 6. Temperature Trend

```{r temperature}
plot(
  weather$Date, weather$Temperature_C,
  type = "o",
  pch = 16,
  xlab = "Date",
  ylab = "Temperature (°C)",
  main = "Daily Temperature Trend"
)
```

**Interpretation:** The graph shows day-to-day changes in temperature. The selected period contains both cooler and warmer days.

# 7. Rainfall Trend

```{r rainfall}
barplot(
  weather$Rainfall_mm,
  names.arg = format(weather$Date, "%d"),
  xlab = "Day of August 2026",
  ylab = "Rainfall (mm)",
  main = "Daily Rainfall Trend",
  las = 2
)
```

**Interpretation:** Rainfall varies considerably. Several days have little or no rainfall, while a few days have substantially higher rainfall.

# 8. Temperature and Rainfall Relationship

```{r scatter}
plot(
  weather$Temperature_C, weather$Rainfall_mm,
  pch = 19,
  xlab = "Temperature (°C)",
  ylab = "Rainfall (mm)",
  main = "Temperature vs Rainfall"
)

abline(lm(Rainfall_mm ~ Temperature_C, data = weather), lty = 2)
```

The Pearson correlation is calculated using:

```{r correlation}
correlation <- cor(weather$Temperature_C, weather$Rainfall_mm)
round(correlation, 3)
```

**Interpretation:** The correlation coefficient measures the linear relationship between temperature and rainfall in this sample. It should not be treated as a general climate conclusion because the dataset covers only 30 days.

# 9. Findings

The analysis identifies:

- Average temperature of approximately **32 °C**.
- Minimum temperature of **28 °C**.
- Maximum temperature of **36 °C**.
- Average daily rainfall of approximately **11.53 mm**.
- Total rainfall of approximately **346 mm**.
- Highest rainfall of **35 mm** during the selected period.
- The correlation value describes the linear relationship between temperature and rainfall in this sample.

# 10. Conclusion

This project demonstrates a complete basic data-analysis workflow in R. The dataset was imported from a CSV file, checked for missing and duplicate records, cleaned, statistically analysed and visualized.

The graphs show that both temperature and rainfall change from day to day. The correlation analysis provides a numerical measure of their linear relationship for the selected sample.

Because this is a small academic dataset covering only 30 days, the findings should not be interpreted as long-term climate behaviour. A larger dataset covering several months or years would be more suitable for long-term weather trend analysis.

# 11. R Functions Used

| Function | Purpose |
|---|---|
| `read.csv()` | Imports CSV data |
| `as.Date()` | Converts text to date format |
| `head()` | Displays first rows |
| `str()` | Shows data structure |
| `is.na()` | Checks missing values |
| `duplicated()` | Checks duplicate rows |
| `complete.cases()` | Identifies complete records |
| `mean()` | Calculates average |
| `median()` | Calculates middle value |
| `min()` / `max()` | Finds minimum / maximum |
| `sd()` | Calculates standard deviation |
| `plot()` | Creates line/scatter plots |
| `barplot()` | Creates bar graph |
| `cor()` | Calculates correlation |
| `lm()` | Fits a linear model |
| `knitr::kable()` | Creates a formatted table |

# 12. Final Result

The project covers **data importing, preprocessing and cleaning, statistical analysis, graph creation, interpretation, R-code explanation, findings and conclusion**, as required for the Tiny Project.
